---
title: "Venlafaxine MaAsLin3 (Multivariate Association with Linear Models) analysis"
author: "Milena Esser"
date: "2025-05-20"
output:
  html_document: 
    toc: true
    toc_depth: 4
    number_sections: true
    toc_float: true
  pdf_document: default
editor_options:
  chunk_output_type: console
---

Different from other differential abundance tools, low prevalence features need not be filtered out since the prevalence modeling in MaAsLin 3 already accounts for high proportions of zeros. However, filtering low prevalence features might improve power.

There are also a few rules of thumb to keep in mind.

-   Models should ideally have about 10 times as many samples (all samples for logistic fits, non-zero samples for linear fits) as covariate terms (all continuous variables plus all categorical variable levels).

-   Coefficients (effect sizes) larger than about 15 in absolute value are usually suspect unless very small unstandardized predictors are being included. (A coefficient of 15 corresponds to a fold change \>30000!). If you encounter such coefficients, check that (1) no error was thrown, (2) the diagnostics look reasonable, (3) a sufficient number of samples were used in fitting, (4) the q-value is significant, (5) the metadata are not highly collinear, and (6) the random effects are plausible.

Maaslin3 Model Selection Strategy\
In Maaslin3, you specify:

-   `fixed_effects`: variables you want to test or control for

-   `random_effects`: grouping variables to account for non-independence (e.g., repeated measures)

# Set up

```{r setup-5, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
rm(list=ls()) #Clears environment
#Change margin size to be smaller so graphs fit in the plot panel
par(mar = c(2, 2, 2, 2)) # Set the margin on all sides to 2
#setwd("C:\\Users\\milen\\OneDrive - McMaster University\\Bow river microbiome\\R_analysis")


###Loading packages
library(phyloseq) #phyloseq will be the main package used for structuring microbiome data and diversity comparisons
library(ggplot2) #For creating graphs
library(ggtext)
library(plyr) #if you load this after dpylr it messes up the dplyr functions
library(dplyr) #Helps with data wrangling 
library(vegan)
library(tidyverse) #data wrangling
library(knitr) #For R Markdown knitting
library(RColorBrewer)
library(ggpubr)
library(picante)
library(ggh4x) # For facet_nested_wrap()
library(multcompView)
library(data.table)

library(maaslin3)
```

# Biofilm

```{r Loading-data-maaslin3-bio}

load("R_analysis/18S/R_output/ps_objects/Venla_biofilm_phyloseq_filt_unnorm.RData")

```

```{r create-maaslin3-output-dir-bio}
# Create output directory
output_dir <- file.path("R_analysis/18S/R_output", "MaAsLin3")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

```

#### Prepare data for analysis

```{r prep-data-for-maaslin3-bio}
#Prepare data for analysis
#Extract abundance (e.g., genus level) Species recommended
ps<- tax_glom(ps_biofilm, taxrank = "Species")

otu <- as.data.frame(otu_table(ps))  # Features
if (taxa_are_rows(ps)) {
  otu <- t(otu)
}

otu <- as.data.frame(otu)
otu <- otu[, sapply(otu, is.numeric)]
# Ensure OTU table is a numeric data frame
otu <- as.data.frame(otu)
otu <- otu[, sapply(otu, is.numeric)]

# Metadata
meta <- as(sample_data(ps), "data.frame")

sapply(meta[, c("treatment_tag", "enclosure")], class)
str(meta[, c("treatment_tag", "enclosure")])

# Drop rows not shared between OTU and meta
shared_ids <- intersect(rownames(otu), rownames(meta))
otu <- otu[shared_ids, , drop = FALSE]
meta <- meta[shared_ids, , drop = FALSE]

# Sort to make sure sample order matches
otu <- otu[order(rownames(otu)), ]
meta <- meta[order(rownames(meta)), ]

# Final sanity check
stopifnot(identical(rownames(otu), rownames(meta)))


# Extract taxonomy table from phyloseq
tax_table_df <- as.data.frame(tax_table(ps))

# Make sure rownames are ASV IDs (should be already)
taxonomy_df <- tax_table_df

# Optional: check taxonomy columns
head(taxonomy_df)
```

```{r rename-otu-table-to-include-species-names-bio}
#create new otu table to have species names in final document:
# First, create the Genus_species label
taxonomy_df$Genus_species <- paste(taxonomy_df$Genus, taxonomy_df$Species, sep = "_")

# Optionally, replace spaces or NAs if present
taxonomy_df$Genus_species[is.na(taxonomy_df$Genus_species)] <- "Unknown_sp"
taxonomy_df$Genus_species <- gsub(" ", "_", taxonomy_df$Genus_species)

# Assign ASV names (rownames in taxonomy_df) to this new column for easy matching
name_map <- taxonomy_df$Genus_species
names(name_map) <- rownames(taxonomy_df)

# Rename OTU table columns using the name map, but only where there is a match
colnames(otu) <- ifelse(colnames(otu) %in% names(name_map),
                        name_map[colnames(otu)],
                        colnames(otu))  # leave unmatched ASVs as-is
otu
```

```{r format-metadata-bio}
# Remove commas, convert to numeric, set control = 0
meta$treatment_conc <- as.numeric(gsub(",", "", as.character(meta$treatment_tag)))
meta$treatment_conc[meta$treatment_tag == "control"] <- 0

# Check it worked
table(meta$treatment_tag, meta$treatment_conc)


```

## MaAsLin3 biofilm

fixed effects: treatment

random effects: enclosure

```{r maaslin3-biofilm-bio}

#create output directory
output_base <- file.path(output_dir, "MaAsLin3_biofilm")

fit_data <- maaslin3(
    input_data = otu,
    input_metadata = meta,
    output = output_base,
    fixed_effects = c("treatment_conc"),
    random_effects = c("enclosure"),
    normalization = 'TSS',
    transform = 'LOG',
    augment = TRUE,  # recommended
    # standardize = TRUE, # uncomment if you want standardized numeric predictors
    max_significance = 0.1,
    median_comparison_abundance = TRUE,
    median_comparison_prevalence = FALSE,
    verbosity = "DEBUG",
    small_random_effects = F
)
```

# Sediment

```{r Loading-data-maaslin3-sed}

# not rarefied dataset "ps_sediment"

load("R_analysis/18S/R_output/ps_objects/Venla_sediment_phyloseq_filt_unnorm.RData")


```

Create output directory

```{r create-maaslin3-output-dir-sed}
# Create output directory
output_dir <- file.path("R_analysis/18S/R_output", "MaAsLin3")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

```

#### Prepare data for analysis

```{r prep-data-for-maaslin3-sed}
#Prepare data for analysis
#Extract abundance (e.g., genus level) Species recommended
ps<- tax_glom(ps_sediment, taxrank = "Species")

otu <- as.data.frame(otu_table(ps))  # Features
if (taxa_are_rows(ps)) {
  otu <- t(otu)
}

otu <- as.data.frame(otu)
otu <- otu[, sapply(otu, is.numeric)]
# Ensure OTU table is a numeric data frame
otu <- as.data.frame(otu)
otu <- otu[, sapply(otu, is.numeric)]

# Metadata
meta <- as(sample_data(ps), "data.frame")

sapply(meta[, c("treatment_tag", "enclosure")], class)
str(meta[, c("treatment_tag", "enclosure")])

# Drop rows not shared between OTU and meta
shared_ids <- intersect(rownames(otu), rownames(meta))
otu <- otu[shared_ids, , drop = FALSE]
meta <- meta[shared_ids, , drop = FALSE]

# Sort to make sure sample order matches
otu <- otu[order(rownames(otu)), ]
meta <- meta[order(rownames(meta)), ]

# Final sanity check
stopifnot(identical(rownames(otu), rownames(meta)))


# Extract taxonomy table from phyloseq
tax_table_df <- as.data.frame(tax_table(ps))

# Make sure rownames are ASV IDs (should be already)
taxonomy_df <- tax_table_df

# Optional: check taxonomy columns
head(taxonomy_df)
```

```{r rename-otu-table-to-include-species-names-sed}
#create new otu table to have species names in final document:
# First, create the Genus_species label
taxonomy_df$Genus_species <- paste(taxonomy_df$Genus, taxonomy_df$Species, sep = "_")

# Optionally, replace spaces or NAs if present
taxonomy_df$Genus_species[is.na(taxonomy_df$Genus_species)] <- "Unknown_sp"
taxonomy_df$Genus_species <- gsub(" ", "_", taxonomy_df$Genus_species)

# Assign ASV names (rownames in taxonomy_df) to this new column for easy matching
name_map <- taxonomy_df$Genus_species
names(name_map) <- rownames(taxonomy_df)

# Rename OTU table columns using the name map, but only where there is a match
colnames(otu) <- ifelse(colnames(otu) %in% names(name_map),
                        name_map[colnames(otu)],
                        colnames(otu))  # leave unmatched ASVs as-is
otu
```

```{r format-metadata-sed}
# Remove commas, convert to numeric, set control = 0
meta$treatment_conc <- as.numeric(gsub(",", "", as.character(meta$treatment_tag)))
meta$treatment_conc[meta$treatment_tag == "control"] <- 0

# Check it worked
table(meta$treatment_tag, meta$treatment_conc)


```

## MaAsLin3 sediment

fixed effects: treatment

random effects: enclosure

```{r maaslin3-sediment}

#create output directory
output_base <- file.path(output_dir, "MaAsLin3_sediment")

fit_data <- maaslin3(
    input_data = otu,
    input_metadata = meta,
    output = output_base,
    fixed_effects = c("treatment_conc"),
    random_effects = c("enclosure"),
    normalization = 'TSS',
    transform = 'LOG',
    augment = TRUE,  # recommended
    # standardize = TRUE, # uncomment if you want standardized numeric predictors
    max_significance = 0.1,
    median_comparison_abundance = TRUE,
    median_comparison_prevalence = FALSE,
    verbosity = "DEBUG",
    small_random_effects = F
)
```
