---
title: "Venlafaxine - 16S beta diversity analysis"
author: "Milena Esser"
date: "2025-05-20"
output:
  html_document: 
    toc: true
    toc_depth: 4
    number_sections: true
    toc_float: true
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r 3-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
rm(list=ls()) #Clears environment
#Change margin size to be smaller so graphs fit in the plot panel
par(mar = c(2, 2, 2, 2)) # Set the margin on all sides to 2
#setwd("C:\\Users\\milen\\OneDrive - McMaster University\\Bow river microbiome\\R_analysis")

###Loading packages
library(phyloseq) #phyloseq will be the main package used for structuring microbiome data and diversity comparisons
library(ggplot2) #For creating graphs
library(ggtext)
library(plyr) #if you load this after dpylr it messes up the dplyr functions
library(dplyr) #Helps with data wrangling 
library(vegan)
library(tidyverse) #data wrangling
library(knitr) #For R Markdown knitting
library(RColorBrewer)
library(ggpubr)
library(MicEco) #to prune based on prevalence
library(picante)
library(ggh4x) # For facet_nested_wrap()
library(multcompView)
library(microbiome) #abundances function Retrieves the taxon abundance table from phyloseq-class object and ensures it is systematically returned as taxa x samples matrix.
library(rdacca.hp) #package for hierarchical and variation partitioning in multiple regression and canonical ananlysis
library(data.table)
```

## Biofilm

```{r Loading data}

# not rarefied dataset "ps_biofilm"
load("R_output/ps_objects/Venla_biofilm_phyloseq_filt_unnorm.RData")

# rarefied dataset "ps_biofilm_rare"
load("R_output/ps_objects/Venla_biofilm_phyloseq_filt_rar.RData")

# css-normalized dataset "ps_biofilm_css"
load("R_output/ps_objects/Venla_biofilm_phyloseq_filt_css.RData")

# TSS-normalized dataset "ps_biofilm_prop"
load("R_output/ps_objects/Venla_biofilm_phyloseq_filt_propnorm.RData")
```

```{r select-data-for-beta-bio}
dataset <- "css"
ps <- get(paste0("ps_biofilm_", dataset))

output_dir <- file.path("R_output", "Beta_diversity", dataset)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

### PCoA Bray Curtis

Performed Principal Coordinates Analysis (PCoA) on the phyloseq object using Bray-Curtis distance. This method reduces multidimensional microbial community data into principal axes to visualize sample similarities.

***PCoA plot***

```{r Plot-PCoA-bio}

bray <- ordinate(ps, method = "PCoA", distance = "bray")

metadata<- data.frame(sample_data(ps))

metadata$enclosure <- factor(metadata$enclosure)
metadata$treatment <- factor(metadata$treatment_tag)


p <- plot_ordination(ps, bray, color = "treatment_tag") +
  theme_classic()

p
#save plot

filename <- file.path(output_dir, paste0("PCoA_Bray_biofilm_", dataset, ".png"))
ggsave(filename = filename, height = 5, width = 7)


```

### NMDS Bray Curtis

I looked at 2D (k=2) and 3D (k=3) NMDS

Interpretation of stress values:

-   \< 0.05 = excellent,

-   0.05–0.1 = very good,

-   0.1–0.2 = usable but some distortion,

-   \> 0.2 = not reliable

Weak ties mean that few identical distances were tied during ranking, which is good, it indicates the data have good variability.

```{r NMDS-all-samples-bio}
# NMDS 
set.seed(501)
# Run NMDS with k=3
ord_NMDS_k3 <- ordinate(ps, method = "NMDS", distance = "bray", trymax = 100, k = 3)
cat("NMDS with k=3 results:\n")
cat("Stress value:", ord_NMDS_k3$stress, "\n")


# Run NMDS with k=2
ord_NMDS_k2 <- ordinate(ps, method = "NMDS", distance = "bray", trymax = 100, k = 2)
cat("NMDS with k=2 results:\n")
cat("Stress value:", ord_NMDS_k2$stress, "\n")

#Weak ties mean that few identical distances were tied during ranking, which is good — it indicates the data have good variability.

```

Based on stress values, I continued with:

```{r select-NMDS-model-bio}

selected_model_name <- "ord_NMDS_k2"
ord_NMDS <- ord_NMDS_k2

cat("Selected model:", selected_model_name, "; Stress value:", ord_NMDS$stress, "\n")
```

```{r plotting-NMDS-bio, fig.height=10, fig.width=14, message=FALSE}
#Plotting all sites and taxa in one facet
plotNMDS1<-plot_ordination(ps, ord_NMDS, color = "treatment_tag") + 
  theme_classic() +
  geom_point(size=3)+
  theme(strip.background = element_blank())+
  theme(legend.key.size = unit(0.3, 'cm'), 
        legend.title = element_text(size=20), 
        legend.text = element_text(size=18), 
        axis.text = element_text(size=16, color="black"),
        axis.title = element_text(size=18, color="black"),
        legend.position = "right")+
  guides(color=guide_legend(title="Treatment")) +
  scale_color_discrete(palette(c("orange","blue","red","black","magenta","peachpuff")))


newtab1 = data.table(plotNMDS1$data)
newtab1$treatment_tag <- ordered(newtab1$treatment_tag,
 levels=c("control", "4", "26", "138", "3,727", "100,000"), labels=c("control", "4", "26", "138", "3,727", "100,000"))
plotNMDS1$data <- newtab1
print(plotNMDS1)

#save plot
filename <- file.path(output_dir, paste0("NMDS_biofilm_", dataset, ".png"))
ggsave(filename = filename, height = 10, width = 14)



```

#### Statistics

##### PERMANOVA

```{r PERMANOVA-all-samples-together-bio}
## Create Bray-Curtis distance matrix
df_bray <- phyloseq::distance(ps, method = "bray")

## PERMANOVA testing for effects of Site and Sample Type on community composition
test.adonis.all <- adonis2(df_bray ~ treatment_tag, data = metadata)

## Print a clean, reader-friendly summary for the report
cat("PERMANOVA Results for Bray-Curtis Distance:\n")
cat("Testing effects of treatment on microbial community composition.\n\n")

# Format the output nicely
adonis_table <- as.data.frame(test.adonis.all)

print(adonis_table)

cat("\nInterpretation:\n")
cat("- Significant p-values (< 0.05) indicate a statistically significant effect of that factor on community composition.\n")
cat("- R² values represent the proportion of variation explained by each factor.\n")
```

##### Beta dispersion

betadisper measures the spread (i.e., variability) of samples around the group centroid in NMDS (or distance) space. It’s analogous to a Levene’s test for homogeneity of variance, but for multivariate data. It helps to understand if your PERMANOVA results are driven more by location (centroid differences) or spread (dispersion differences).

If dispersions differ too much, you have to be cautious in interpreting PERMANOVA results. As this is highly significant: Differences in microbial composition among Families are not only due to changes in the mean community but also because some Families are more variable than others.

-   **If p \> 0.05**: No evidence of unequal dispersion → You can trust group separation in PERMANOVA.
-   **If p \< 0.05**: Significant dispersion → Group differences in PERMANOVA **may be confounded** by group variance, **not just location**.

```{r beta-dispersion-all-samples-together-bio}

#Are microbial (or community) compositions more variable in some Families than in others?
bd_all<-betadisper(df_bray, metadata$treatment_tag) 
anova(bd_all)


#### Boxplot to show dispersion #####
# Extract distances to centroids
beta_df <- data.frame(
  treatment = bd_all$group,
  DistanceToCentroid = bd_all$distances
)

# Show first few rows to verify structure
cat("Summary of distances to group centroids:\n")
print(head(beta_df))

```

```{r tukey-to-identify-sig-diff-dispersion-bio, eval=F}

#In case of p< 0.05: pairwise comparison to identify group with significantly different dispersion

tukey_res <- TukeyHSD(bd_all_fam)
sig_disp <- as.data.frame(tukey_res$group)
sig_disp$Comparison <- rownames(sig_disp)
sig_disp <- sig_disp[sig_disp$`p adj` < 0.05, ]

# View the significant ones
cat("All significant results:\n")
sig_disp[order(sig_disp$`p adj`), ]
```

Create boxplot

```{r beta-dispersion-boxplot-all-bio, fig.height=5, fig.width=8}
# Create a ggplot boxplot
ggplot(beta_df, aes(x = treatment, y = DistanceToCentroid, fill = treatment)) +
  geom_boxplot(alpha = 0.7, outlier.color = "black", outlier.size = 1.5) +
  geom_jitter(width = 0.2, alpha=0.5, size=1.5) + # add jittered points
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size=12),
    axis.text.y = element_text(size=12),
    axis.title = element_text(size=14, face="bold"),
    legend.position = "none"
  ) +
  labs(
    x = "treatment",
    y = "Distance to Group Centroid",
 #   title = "Beta Dispersion across Invertebrate Families"
  ) +
  scale_fill_brewer(palette = "Set2")  # optional color palette

#Save plots

#Save plot as png
filename <- file.path(output_dir, paste0("Beta_Dispersion_across_treatmentgroups_", dataset, ".png"))
ggsave(filename = filename, height = 5, width = 8)
#save plot as pdf
filename <- file.path(output_dir, paste0("Beta_Dispersion_across_treatmentgroups_", dataset, ".pdf"))
ggsave(filename = filename, height = 5, width = 8)


#Beta dispersion differed significantly among Families, suggesting that observed differences in community composition are associated with both shifts in mean composition and differences in within-group variability.
```

## Sediment

```{r Loading data-sed}

# not rarefied dataset "ps_biofilm"
load("R_output/ps_objects/Venla_sediment_phyloseq_filt_unnorm.RData")

# rarefied dataset "ps_biofilm_rare"
load("R_output/ps_objects/Venla_sediment_phyloseq_filt_rar.RData")

# css-normalized dataset "ps_biofilm_css"
load("R_output/ps_objects/Venla_sediment_phyloseq_filt_css.RData")

# TSS-normalized dataset "ps_biofilm_prop"
load("R_output/ps_objects/Venla_sediment_phyloseq_filt_propnorm.RData")
```

```{r select-data-for-beta-sed}
dataset <- "css"
ps <- get(paste0("ps_sediment_", dataset))

output_dir <- file.path("R_output", "Beta_diversity", dataset)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

### PCoA Bray Curtis

Performed Principal Coordinates Analysis (PCoA) on the phyloseq object using Bray-Curtis distance. This method reduces multidimensional microbial community data into principal axes to visualize sample similarities.

***PCoA plot***

```{r Plot-PCoA-sed}

bray <- ordinate(ps, method = "PCoA", distance = "bray")

metadata<- data.frame(sample_data(ps))

metadata$enclosure <- factor(metadata$enclosure)
metadata$treatment <- factor(metadata$treatment_tag)


p <- plot_ordination(ps, bray, color = "treatment_tag") +
  theme_classic()

p
#save plot

filename <- file.path(output_dir, paste0("PCoA_Bray_sediment_", dataset, ".png"))
ggsave(filename = filename, height = 5, width = 7)


```

### NMDS Bray Curtis

I looked at 2D (k=2) and 3D (k=3) NMDS

Interpretation of stress values:

-   \< 0.05 = excellent,

-   0.05–0.1 = very good,

-   0.1–0.2 = usable but some distortion,

-   \> 0.2 = not reliable

Weak ties mean that few identical distances were tied during ranking, which is good, it indicates the data have good variability.

```{r NMDS-all-samples-sed}
# NMDS 
set.seed(501)
# Run NMDS with k=3
ord_NMDS_k3 <- ordinate(ps, method = "NMDS", distance = "bray", trymax = 100, k = 3)
cat("NMDS with k=3 results:\n")
cat("Stress value:", ord_NMDS_k3$stress, "\n")


# Run NMDS with k=2
ord_NMDS_k2 <- ordinate(ps, method = "NMDS", distance = "bray", trymax = 100, k = 2)
cat("NMDS with k=2 results:\n")
cat("Stress value:", ord_NMDS_k2$stress, "\n")

#Weak ties mean that few identical distances were tied during ranking, which is good — it indicates the data have good variability.

```

Based on stress values, I continued with:

```{r select-NMDS-model}

selected_model_name <- "ord_NMDS_k2"
ord_NMDS <- ord_NMDS_k2

cat("Selected model:", selected_model_name, "; Stress value:", ord_NMDS$stress, "\n")
```

```{r plotting-NMDS, fig.height=10, fig.width=14, message=FALSE}
#Plotting all sites and taxa in one facet
plotNMDS1<-plot_ordination(ps, ord_NMDS, color = "treatment_tag") + 
  theme_classic() +
  geom_point(size=3)+
  theme(strip.background = element_blank())+
  theme(legend.key.size = unit(0.3, 'cm'), 
        legend.title = element_text(size=20), 
        legend.text = element_text(size=18), 
        axis.text = element_text(size=16, color="black"),
        axis.title = element_text(size=18, color="black"),
        legend.position = "right")+
  guides(color=guide_legend(title="Treatment")) +
  scale_color_discrete(palette(c("orange","blue","red","black","magenta","peachpuff")))


newtab1 = data.table(plotNMDS1$data)
newtab1$treatment_tag <- ordered(newtab1$treatment_tag,
 levels=c("control", "4", "26", "138", "3,727", "100,000"), labels=c("control", "4", "26", "138", "3,727", "100,000"))
plotNMDS1$data <- newtab1
print(plotNMDS1)

#save plot
filename <- file.path(output_dir, paste0("NMDS_sediment_", dataset, ".png"))
ggsave(filename = filename, height = 10, width = 14)



```

#### Statistics

##### PERMANOVA

```{r PERMANOVA-all-samples-together-sed}
## Create Bray-Curtis distance matrix
df_bray <- phyloseq::distance(ps, method = "bray")

## PERMANOVA testing for effects of Site and Sample Type on community composition
test.adonis.all <- adonis2(df_bray ~ treatment_tag, data = metadata)

## Print a clean, reader-friendly summary for the report
cat("PERMANOVA Results for Bray-Curtis Distance:\n")
cat("Testing effects of treatment on microbial community composition.\n\n")

# Format the output nicely
adonis_table <- as.data.frame(test.adonis.all)

print(adonis_table)

cat("\nInterpretation:\n")
cat("- Significant p-values (< 0.05) indicate a statistically significant effect of that factor on community composition.\n")
cat("- R² values represent the proportion of variation explained by each factor.\n")
```

##### Beta dispersion

betadisper measures the spread (i.e., variability) of samples around the group centroid in NMDS (or distance) space. It’s analogous to a Levene’s test for homogeneity of variance, but for multivariate data. It helps to understand if your PERMANOVA results are driven more by location (centroid differences) or spread (dispersion differences).

If dispersions differ too much, you have to be cautious in interpreting PERMANOVA results. As this is highly significant: Differences in microbial composition among Families are not only due to changes in the mean community but also because some Families are more variable than others.

-   **If p \> 0.05**: No evidence of unequal dispersion → You can trust group separation in PERMANOVA.
-   **If p \< 0.05**: Significant dispersion → Group differences in PERMANOVA **may be confounded** by group variance, **not just location**.

```{r beta-dispersion-all-samples-together-sed}

#Are microbial (or community) compositions more variable in some Families than in others?
bd_all<-betadisper(df_bray, metadata$treatment_tag) 
anova(bd_all)


#### Boxplot to show dispersion #####
# Extract distances to centroids
beta_df <- data.frame(
  treatment = bd_all$group,
  DistanceToCentroid = bd_all$distances
)

# Show first few rows to verify structure
cat("Summary of distances to group centroids:\n")
print(head(beta_df))

```

```{r tukey-to-identify-sig-diff-dispersion, eval=F}

#In case of p< 0.05: pairwise comparison to identify group with significantly different dispersion

tukey_res <- TukeyHSD(bd_all_fam)
sig_disp <- as.data.frame(tukey_res$group)
sig_disp$Comparison <- rownames(sig_disp)
sig_disp <- sig_disp[sig_disp$`p adj` < 0.05, ]

# View the significant ones
cat("All significant results:\n")
sig_disp[order(sig_disp$`p adj`), ]
```

Create boxplot

```{r beta-dispersion-boxplot-all-sed, fig.height=5, fig.width=8}
# Create a ggplot boxplot
ggplot(beta_df, aes(x = treatment, y = DistanceToCentroid, fill = treatment)) +
  geom_boxplot(alpha = 0.7, outlier.color = "black", outlier.size = 1.5) +
  geom_jitter(width = 0.2, alpha=0.5, size=1.5) + # add jittered points
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size=12),
    axis.text.y = element_text(size=12),
    axis.title = element_text(size=14, face="bold"),
    legend.position = "none"
  ) +
  labs(
    x = "treatment",
    y = "Distance to Group Centroid",
 #   title = "Beta Dispersion across Invertebrate Families"
  ) +
  scale_fill_brewer(palette = "Set2")  # optional color palette

#Save plots

#Save plot as png
filename <- file.path(output_dir, paste0("Beta_Dispersion_across_treatmentgroups_sediment", dataset, ".png"))
ggsave(filename = filename, height = 5, width = 8)
#save plot as pdf
filename <- file.path(output_dir, paste0("Beta_Dispersion_across_treatmentgroups_sediment", dataset, ".pdf"))
ggsave(filename = filename, height = 5, width = 8)


#Beta dispersion differed significantly among Families, suggesting that observed differences in community composition are associated with both shifts in mean composition and differences in within-group variability.
```

## Water

```{r Loading data-water}

# not rarefied dataset "ps_biofilm"
load("R_output/ps_objects/Venla_water_phyloseq_filt_unnorm.RData")

# rarefied dataset "ps_biofilm_rare"
load("R_output/ps_objects/Venla_water_phyloseq_filt_rar.RData")

# css-normalized dataset "ps_biofilm_css"
load("R_output/ps_objects/Venla_water_phyloseq_filt_css.RData")

# TSS-normalized dataset "ps_biofilm_prop"
load("R_output/ps_objects/Venla_water_phyloseq_filt_propnorm.RData")
```

```{r select-data-for-beta_wat}
dataset <- "css"
ps_water <- get(paste0("ps_water_", dataset))

# Remove samples with "BLANK" in enclosure
ps <- subset_samples(ps_water, !grepl("BLANK", enclosure, ignore.case = TRUE))


output_dir <- file.path("R_output", "Beta_diversity", dataset)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

### PCoA Bray Curtis

Performed Principal Coordinates Analysis (PCoA) on the phyloseq object using Bray-Curtis distance. This method reduces multidimensional microbial community data into principal axes to visualize sample similarities.

***PCoA plot***

```{r Plot-PCoA-water}

bray <- ordinate(ps, method = "PCoA", distance = "bray")

metadata<- data.frame(sample_data(ps))

metadata$enclosure <- factor(metadata$enclosure)
metadata$treatment <- factor(metadata$treatment_tag)


p <- plot_ordination(ps, bray, color = "treatment_tag") +
  theme_classic()

p
#save plot

filename <- file.path(output_dir, paste0("PCoA_Bray_water_", dataset, ".png"))
ggsave(filename = filename, height = 5, width = 7)


```

### NMDS Bray Curtis

I looked at 2D (k=2) and 3D (k=3) NMDS

Interpretation of stress values:

-   \< 0.05 = excellent,

-   0.05–0.1 = very good,

-   0.1–0.2 = usable but some distortion,

-   \> 0.2 = not reliable

Weak ties mean that few identical distances were tied during ranking, which is good, it indicates the data have good variability.

```{r NMDS-all-samples-water}
# NMDS 
set.seed(501)
# Run NMDS with k=3
ord_NMDS_k3 <- ordinate(ps, method = "NMDS", distance = "bray", trymax = 100, k = 3)
cat("NMDS with k=3 results:\n")
cat("Stress value:", ord_NMDS_k3$stress, "\n")


# Run NMDS with k=2
ord_NMDS_k2 <- ordinate(ps, method = "NMDS", distance = "bray", trymax = 100, k = 2)
cat("NMDS with k=2 results:\n")
cat("Stress value:", ord_NMDS_k2$stress, "\n")

#Weak ties mean that few identical distances were tied during ranking, which is good — it indicates the data have good variability.

```

Based on stress values, I continued with:

```{r select-NMDS-model-water}

selected_model_name <- "ord_NMDS_k2"
ord_NMDS <- ord_NMDS_k2

cat("Selected model:", selected_model_name, "; Stress value:", ord_NMDS$stress, "\n")
```

```{r plotting-NMDS-water, fig.height=10, fig.width=14, message=FALSE}
#Plotting all sites and taxa in one facet
plotNMDS1<-plot_ordination(ps, ord_NMDS, color = "treatment_tag") + 
  theme_classic() +
  geom_point(size=3)+
  theme(strip.background = element_blank())+
  theme(legend.key.size = unit(0.3, 'cm'), 
        legend.title = element_text(size=20), 
        legend.text = element_text(size=18), 
        axis.text = element_text(size=16, color="black"),
        axis.title = element_text(size=18, color="black"),
        legend.position = "right")+
  guides(color=guide_legend(title="Treatment")) +
  scale_color_discrete(palette(c("orange","blue","red","black","magenta","peachpuff")))


newtab1 = data.table(plotNMDS1$data)
newtab1$treatment_tag <- ordered(newtab1$treatment_tag,
 levels=c("control", "4", "26", "138", "3,727", "100,000"), labels=c("control", "4", "26", "138", "3,727", "100,000"))
plotNMDS1$data <- newtab1
print(plotNMDS1)

#save plot
filename <- file.path(output_dir, paste0("NMDS_water_", dataset, ".png"))
ggsave(filename = filename, height = 10, width = 14)



```

#### Statistics

##### PERMANOVA

```{r PERMANOVA-all-samples-together-water}
## Create Bray-Curtis distance matrix
df_bray <- phyloseq::distance(ps, method = "bray")

## PERMANOVA testing for effects of Site and Sample Type on community composition
test.adonis.all <- adonis2(df_bray ~ treatment_tag, data = metadata)

## Print a clean, reader-friendly summary for the report
cat("PERMANOVA Results for Bray-Curtis Distance:\n")
cat("Testing effects of treatment on microbial community composition.\n\n")

# Format the output nicely
adonis_table <- as.data.frame(test.adonis.all)

print(adonis_table)

cat("\nInterpretation:\n")
cat("- Significant p-values (< 0.05) indicate a statistically significant effect of that factor on community composition.\n")
cat("- R² values represent the proportion of variation explained by each factor.\n")
```

##### Beta dispersion

betadisper measures the spread (i.e., variability) of samples around the group centroid in NMDS (or distance) space. It’s analogous to a Levene’s test for homogeneity of variance, but for multivariate data. It helps to understand if your PERMANOVA results are driven more by location (centroid differences) or spread (dispersion differences).

If dispersions differ too much, you have to be cautious in interpreting PERMANOVA results. As this is highly significant: Differences in microbial composition among Families are not only due to changes in the mean community but also because some Families are more variable than others.

-   **If p \> 0.05**: No evidence of unequal dispersion → You can trust group separation in PERMANOVA.
-   **If p \< 0.05**: Significant dispersion → Group differences in PERMANOVA **may be confounded** by group variance, **not just location**.

```{r beta-dispersion-all-samples-together-water}

#Are microbial (or community) compositions more variable in some Families than in others?
bd_all<-betadisper(df_bray, metadata$treatment_tag) 
anova(bd_all)


#### Boxplot to show dispersion #####
# Extract distances to centroids
beta_df <- data.frame(
  treatment = bd_all$group,
  DistanceToCentroid = bd_all$distances
)

# Show first few rows to verify structure
cat("Summary of distances to group centroids:\n")
print(head(beta_df))

```

```{r tukey-to-identify-sig-diff-dispersion-water, eval=F}

#In case of p< 0.05: pairwise comparison to identify group with significantly different dispersion

tukey_res <- TukeyHSD(bd_all_fam)
sig_disp <- as.data.frame(tukey_res$group)
sig_disp$Comparison <- rownames(sig_disp)
sig_disp <- sig_disp[sig_disp$`p adj` < 0.05, ]

# View the significant ones
cat("All significant results:\n")
sig_disp[order(sig_disp$`p adj`), ]
```

Create boxplot

```{r beta-dispersion-boxplot-all-water, fig.height=5, fig.width=8}
# Create a ggplot boxplot
ggplot(beta_df, aes(x = treatment, y = DistanceToCentroid, fill = treatment)) +
  geom_boxplot(alpha = 0.7, outlier.color = "black", outlier.size = 1.5) +
  geom_jitter(width = 0.2, alpha=0.5, size=1.5) + # add jittered points
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size=12),
    axis.text.y = element_text(size=12),
    axis.title = element_text(size=14, face="bold"),
    legend.position = "none"
  ) +
  labs(
    x = "treatment",
    y = "Distance to Group Centroid",
 #   title = "Beta Dispersion across Invertebrate Families"
  ) +
  scale_fill_brewer(palette = "Set2")  # optional color palette

#Save plots

#Save plot as png
filename <- file.path(output_dir, paste0("Beta_Dispersion_across_treatmentgroups_water_", dataset, ".png"))
ggsave(filename = filename, height = 5, width = 8)
#save plot as pdf
filename <- file.path(output_dir, paste0("Beta_Dispersion_across_treatmentgroups_water_", dataset, ".pdf"))
ggsave(filename = filename, height = 5, width = 8)


#Beta dispersion differed significantly among Families, suggesting that observed differences in community composition are associated with both shifts in mean composition and differences in within-group variability.
```
